---
title: "Homework 3"
author: "Dan Schumacher, hdd249"
format: html
editor: visual
---

```{r}
#| include: False
library(ggplot2)
library(dplyr)
library(MASS)
library(car)
library(olsrr)
```

# Algorithms Homework 3

## Exercise 1:

The medical director at your company wants to know if Weight alone can predict Cholesterol outcomes. Consider modeling Cholesterol as a function of Weight.

### a)

Fit a linear regression model for Cholesterol as a function of Weight. If any points are unduly influential, note those points, then remove them and refit the model. Consider Cook's distance cut-off to be 0.015.

*Unduly influential points are point 23 and points 210 (shout out to San Antonio's area code!)*

```{r}
#set up data
df_heart <- 
  read.csv('./data/heart.csv')

# fit linear model
lm_heart <- 
  lm(
    formula = Cholesterol ~ Weight,
     data = df_heart
    )

#diagnostics plot
par(mfrow = c(2,2))
plot(lm_heart, which = c(1:4))

# find influential points
inf.id <-
  which(
    cooks.distance(lm_heart)>0.015
    )

#print unduly influential points
print(
  df_heart[inf.id,]
  )

#remove those points and refit model
lm_heart_rm <-
  lm(
    formula = Cholesterol ~ Weight,
    data = df_heart[-inf.id,]
  )
```

### b)

Comment on the significance of the parameters

-   Our overall p is less than .05 and our Weight-p is less than .05. Both are significant.

```{r}
summary(lm_heart_rm)
```

-   variation explained by the model

    -   with an ${R}^2$ (rounded to) .006, our model can only explain .6% of the variation.

-   any remaining issues noted in the diagnostics plots

```         
-   **Residuals vs Fitted plot**: we want it to be random and centered roughly around 0. Looks good to me

-   **Q-Q plot**: it follows the line pretty well! Although there is a slight curve toward the end, I don't think it is steep enough to discount our findings

-   **Scale-Location**: Most of our residuals fall below 1.5 which supports normality. However a good number also fall above that line. This could be a concern.

-   **Cook's Distance**: we fixed this already
```

```{r}
par(mfrow = c(2,2))
plot(lm_heart_rm, which = c(1:4))
```

What does this model tell us about the relationship between Cholesterol and Weight? Interpret the relationship specifically.

-   Our model tells us that for every unit increase in weight, cholesterol goes up by .123.

-   Cholesterol = .123(weight) + 203.58

Explain to the medical director whether this is a good model for predicting Cholesterol levels.

-   While our model is significant (low p-val) and proves that there is a linear relationship between Cholesterol and Weight, with such a low ${R}^2$, I think that there are better models out there especially if the director is interested in predictive power!

## Exercise 2:

The medical director wants to know if blood pressure and weight can better predict cholesterol outcomes. Consider modeling cholesterol as a function of diastolic, systolic, and weight.

### a)

Fit a linear regression model for cholesterol as a function of diastolic, systolic, and weight.

```{r}
# fit linear model
lm_heart_mult <- 
  lm(
    formula = Cholesterol ~ .,
     data = df_heart
    )
```

Generate the diagnostics plots and comment on any issues that need to be noted. For Cook's distances, do not leave any points that have Cook's distance greater than 0.015.

-   **QQ Plot**: There is a slight upward bend like before. but I don't think it is an issue.

```{r}
# find influential points
inf.id <-
  which(
    cooks.distance(lm_heart_mult) > 0.015
    )

#print unduly influential points
print(
  df_heart[inf.id,]
  )

#remove those points and refit model
lm_heart_mult_rm <-
  lm(
    formula = Cholesterol ~ .,
    data = df_heart[-inf.id, ]
  )

#diagnostics plots
par(mfrow = c(2,2))
plot(lm_heart_mult_rm, which = c(1:4))
```

### b)

```{r}
summary(lm_heart_mult_rm)
```

Comment on the significance of the parameters and how much variation in cholesterol is described by the model.

-   The model as a whole is significant. Diastolic and Systolic are significant.

-   We now have and ${R}^2$ of .038; our model can explain 3.8% of the variance in cholesterol.

Comment on the relationship between cholesterol and statistically significant predictor(s).

-   TO DO THIS WOULD WE NEED TO MAKE A NEW MODEL?

-   Cholesterol = 156.33 + 0.30(Sys) + .25(Dias)

Check multicollinearity issues among predictors.

-   no issues (I'm surprised! I thought that Diastolic and Systolic would be *for sure* correlated)

```{r}
#if VIF > 10 , 90% of xj can be explained by another statistic.
# none of our vars are > 10 
vif(lm_heart_mult_rm)
```

Explain to the medical director whether this is a good model for predicting Cholesterol levels.

-   This is a much better model for predicting Cholesterol levels than the previous model. Now our model can explain 3.8% of the variance (more than 6x more variance explained). Because Weight is not significant, I would be interested in making one more model that dropped weight before presenting it to the medical director. If it performed worse than this current model, I would argue that while our model isn't good at predicting cholesterol levels, it is still useful to see the relationship between these variables.

## Exercise 3:

Now consider step wise model selection for the Cholesterol model. Before performing the model selection, we remove influential points detected in Exercise 2, which have a cook's distance larger than 0.015.

### a)

```{r}
#here is our data without the influential points
df_heart_rm <- 
  df_heart[-inf.id, ]
```

Perform step wise model selection with .05 criteria and address any issues in diagnostics plots.

-   I don't see any issues in the diagnostics plots

```{r}
# stepwise selection
model.stepwise <- ols_step_both_p(lm_heart_mult_rm, pent = .05, prem = .05, details = F)
print(model.stepwise)
plot(model.stepwise)

# reform model based off of ^^^
step_model <-
  lm(Cholesterol ~ Systolic + Diastolic, data = df_heart_rm)

# diagnostics plot
par(mfrow = c(2,2))
plot(lm_heart_mult_rm, which = c(1:4))
```

### b)

Interpret the final model and comment on the variation in Cholesterol explained. Compare the variations explained by the models from Exercises 1 and 2.

-   Our final model has an ${R}^2$ of .0358; this means that our model can explain 3.58% of the variance in Cholesterol. This is, again, about six times the amount of explained variation compared to Ex. 1, but compared to Ex. 2 it is a little bit less. However, all of our features this time are significant which, although it weakens the predictive power of our model, helps interpret-ability.

## Exercise 4:

Now consider the best subset selection for the Cholesterol model. Again, we remove influential points detected in Exercise 2, which has a cook's distance larger than 0.015, before performing the model selection.

### a)

```{r}
# stepwise selection
model.best.subset <- ols_step_best_subset(lm_heart_mult_rm)
model.best.subset

# reform model based off of ^^^
# step_model <-
#   lm(Cholesterol ~ Systolic + Diastolic, data = df_heart_rm)
# 
# # diagnostics plot
# par(mfrow = c(2,2))
# plot(lm_heart_mult_rm, which = c(1:4))
```

Find the best model based on adjusted-R square criteria and specify which predictors are selected.

-   Model 2 (Diastolic Systolic)

### b)

Find the best model based on AIC criteria and specify which predictors are selected.

-   Model 2 (Diastolic Systolic)

### c)

Compare the final models selected in a) and b).

-   They are the same!

Also, compare the final models from the best subset approach with the final model from the step wise selection.

-   They are the same!\
